\MakeUppercase{Graph Product Structure for Non-Minor-Closed Classes}

### 2022-03-12 k-planar.tex
Tidying up

Our proof of \cref{1-planar} follows quickly from the following technical lemma, which is an extension of the analogous result for plane graphs \cite{DJMMUW20}.
\begin{lem}
	\label{induction} The setup:
	\begin{compactenum}
		\item Let $G$ and $G'$ be defined as above.
		\item Let $T$ be a BFS spanning tree of $G'$ rooted at some vertex $r$.
		\item For every integer $j\ge 0$, let $L_j=\{v\in V(G):\dist_T(r,v)=j\}$.
		\item Let $F$ be a cycle in $G'$ with $r$ in the exterior of $F$ and such that $V(F)$ can be partitioned into $k\le 3$ sets $P_1,\ldots,P_k$ such that for each $i\in\{1,\ldots,k\}$,
		\begin{compactenum}
			\item $F[P_i]$ is a path; and
			\item $|V(P_i)\cap L_j| \le 3$ for every integer $j\ge 0$.
		\end{compactenum}
		\item Let $N$ and $N'$ be the subgraphs of $G$ and $G'$ consisting only of those edges and vertices contained in the closure of $F$.
	\end{compactenum}
	Then $N$ has an $H$-partition $\PP=\{S_x\cup\{v_x\} : x\in V(H)\}$ such that:
	\begin{compactenum}[(i)]
		\item $H$ is planar;
		\item for every integer $j\ge 0$ and every $x\in V(H)$, $|S_x\cap L_j|\le 3$;
		\item for each $i\in\{1,\ldots,k\}$, there exists $x_i\in V(H)$ such that $P_i=S_{x_i}$;
		\item $H$ has a tree-decomposition in which every bag has size at most 4 and such that some bag contains $x_1,\ldots,x_k$.
	\end{compactenum}
\end{lem}

@@

  The proof is by induction on the number of inner faces of $N'$.  If $N'$ has only one inner face then we can simply take $\PP:=\{P_1,\ldots,P_k\}$ and verify that the preconditions of the lemma (the `setup') ensure that $\PP$ satisfies the requirements (i)--(iv):
  \begin{inparaenum}[(i)]
    \item $H:=N/\mathcal{P}$ is planar because any graph on $k\le 3$ vertices is planar.
    \item Assumption 4(a) ensures that $|S_x\cap L_i|\le 3$ for each $x\in V(H)$.
    \item By definition, each each of $P_1,\ldots,P_k$ is s part of $\mathcal{P}$.
    \item The trivial tree decomposition of $H$ that places $x_1,\ldots,x_k$ in one bag satisfies this requirement.
  \end{inparaenum}
  Therefore, we may assume that $N$ has more than one inner face.

  The $H$-partition $\mathcal{P}$ will include $P_1$, $P_2$ , $P_3$ and a part $Y$, called a \emph{tripod}, that we now define.  The tripod $Y$ is determined by three vertices $v_1v_2v_3$ that belong to a common inner face $\tau$ of $N'$.  Depending on the value of $k$, there are two cases to consider:
  \begin{compactenum}
    \item If $k\in\{1,2\}$ then we let $\tau:=v_1,\ldots,v_p$ be an inner face of $N'$ that contains at least one edge of $F$ on its boundary.

@@

  We can now define the vertices in the tripod $Y$.  For each $i\in\{1,2,3\}$, let $Q_i$ be the shortest path, in $T$, from $v_i$ to a vertex in $V(F)$.  Let $\overline{Y}$ denote the subgraph of $N'$ consisting of vertices and edges of $Q_1$, $Q_2$, $Q_3$, and $\tau$.  Let $S:=V(Q_1)\cup V(Q_2)\cup V(Q_3)$ and let $v:=v_p$.  Observe that, for each $i\in\{1,2,3\}$ and each $j\ge 0$, $|V(Q_i)\cap L_j|\le 1$.  Therefore $|S\cap L_j|\le 3$ for each integer $j\ge 0$.  Therefore, taking $Y:=S\cup\{v\}\setminus V(F)$ as a part in the partition $\PP$ will not violate condition~(ii).

@@

  Let $F_1,\ldots,F_m,\tau$ be the inner faces of $M$. By our choice of $\tau$, the graph $M$ is $2$-connected.  Furthermore, $V(F_i)$ can be partitioned into at most three sets $P_{i,1}$, $P_{i,2}$, and $P_{i,3}$ where $P_{i,1}\subset Y$, $P_{i,2}\subseteq P_b$, and $P_{i,3}\subseteq P_c$ for some $b,c\in\{1,\ldots,k\}$.  In the following, in order to avoid introducing even more notation or constantly adding the modifier ``when it exists'', we will assume that $V(F_i)\cap Y\neq\emptyset$ and that there are exactly two distinct $b,c\in\{1,\ldots,k\}$ such that $V(F_i)\cap P_b\neq\emptyset$ and $V(F_i)\cap P_c\neq\emptyset$.  When this assumption does not hold, exactly the same reasoning can be used, while omitting one or more of $P_{i,1}$, $P_{i,2}$ or $P_{i,3}$.

  Since $M$ is $2$-connected and $\overline{Y}$ is connected, $F_i[P_{i,j}]$ is a path, for each $j\in\{1,2,3\}$.  Let $N_i$ and $N_i'$ be the subgraphs of $N$ and $N'$ contained in the closure of $F_i$.  Every inner face of $N_i'$ is an inner face of $N'$ and $\tau$ is not an inner face of $N_i'$.  Therefore $N_i'$ has fewer inner faces than $N'$.  Therefore, we can apply induction using the cycle $F_i$ and the partition $P_{i,1},P_{i,2},P_{i,3}$ of $V(F_i)$.  The result of the induction is a $H_i$-partition $\mathcal{P}_i$ of the subgraph $N_i$ of $N$ contained in the closure of $F_i$.
  % The partition $\mathcal{P}_i$ contains each of $P_{i,1}$, $P_{i,2}$ and $P_{i,3}$.
  We can now define our desired partition
  \[
      \mathcal{P}:=\{Y, P_1,\ldots,P_k\} \cup \bigcup_{i=1}^m \{P\in\mathcal{P}_i: P\cap V(M)=\emptyset\} \enspace .
  \]
  It is straightforward to verify that $\mathcal{P}$ is indeed a partition of $V(N)$ and, by definition this is an $H$-partition of $N$ for the graph $H:=N/\mathcal{P}$.  All that remains is to verify that $\mathcal{P}$ satisfies conditions~(i)--(iv).
  \begin{compactenum}[(i)]
    \item To see that $H$ is planar, observe that $N/\mathcal{P}=N'/\mathcal{P}$\todo{Explain why $N/\mathcal{P}=N'/\mathcal{P}$} and that each part $P\in\mathcal{P}$ induces a connected subgraph $G'[P]$ of $G'$.  Since $G'$ is planar and planarity is preserved under edge contractions, this implies that $G'/\mathcal{P}$ is planar.

    \item  By Assumption~4(b) none of the parts $P_1,P_2,P_3$ violate condition~(ii).  We have already argued that the part $Y$ does not violate condition~(ii).  The inductive hypothesis ensures that each of the remaining parts satisfies condition~(ii).

    \item Condition~(iii) is satisfied since $\mathcal{P}$ contains parts $P_1,\ldots,P_k$.

@@

    The tree $T_0$ has a node $y_0$ with $B_{y_0}:=\{x_0,\ldots,x_k\}$.
    % For each $i\in\{1,\ldots,m\}$, let $N_i$ denote the subgraph of $N$ contained in $F_i$.
    The inductive hypothesis ensures that $H_i:=N_i/\mathcal{P}_i$ has a tree decomposition $(B_z:z\in V(T_i))$ whose largest bag has size at most $4$ and that has some bag $B_{z_i}$ that contains the vertices $a_i,b_i,c_i\in V(H_i)$ obtained by contracting $P_{i,1}$, $P_{i,2}$, and $P_{i,3}$, respectively.   Recall that $P_{i,1}:=V(N_i)\cap Y$, $P_{i,2}:=V(N_i)\cap P_b$ and $P_{i,3}:=V(N_i)\cap P_c$ for some $b,c\in\{1,\ldots,k\}$.  For each bag $B_z$, $z\in V(T_i)$, replace each occurrence of $a_i$ by $x_0$, replace each occurrence of $b_i$ by $x_b$ and replace each occurrence of $c_i$ by $x_c$.  These replacements do not increase the size of any bag and they result in a tree decomposition of the graph obtained from $H_i$ by renaming the vertices $a_i$, $b_i$, and $c_i$ with the names $x_0$, $x_a$, and $x_b$. Now add an edge that joins $y_0$ to $z_i$.

    Using the fact that each component of $G-V(M)$ is contained in a single face of $M$, it is now straightforward to verify that, by doing this for each $i\in\{1,\ldots,m\}$ we obtain a tree decomposition of $N$ that satisfies condition~(iv).  \qedhere
  \end{compactenum}

@@

	Let $G'$ be the plane graph obtained by removing each pair of crossing edges from $\overline{G}$ and let $T$ be a BFS tree of $G'$ rooted at $r$.  Now apply \cref{induction} with $G=\overline{G}$, $G'$, $F$, and $P_i=\{v_i\}$ for each $i\in\{1,2,3\}$.  This gives an $H$-partition $\mathcal{P}:=\{S_x\cup\{v_x\}:x\in V(H)\}$ of $\overline{G}-\{r\}\supseteq G$ in which $H$ is planar and has treewidth at most 3.

	Use the layering $\mathcal{L}=\langle L_0',L_1'\ldots\rangle$ where $L_i'=L_{2i}\cup L_{2i+1}$ for each integer $i\ge 0$. That this is a layering of $G$ follows from the fact that $\dist_{G'}(v,w)\le 2$ for each edge $vw\in E(G)$.  Since $|L_j\cap S_x|\le 3$ for each integer $j\ge 0$, $|L_i'\cap (S_x\cup\{v_x\})|\le 7$ for each integer $i\ge 0$ and each $x\in V(H)$. The result follows from \cref{PartitionProduct}.
\end{proof}

>>>>

Our proof of \cref{1-planar} follows quickly from the following technical lemma, which is an extension of the analogous result for plane graphs \cite{DJMMUW20}.
\begin{lem}
	\label{induction} The setup:
	\begin{compactenum}
		\item Let $G$ and $G'$ be defined as above.
		\item Let $T$ be a BFS spanning tree of $G'$ rooted at some vertex $r$.
		\item For every integer $j\ge 0$, let $L_j=\{v\in V(G):\dist_T(r,v)=j\}$.
		\item Let $F$ be a cycle in $G'$ with $r$ in the exterior of $F$ and such that $V(F)$ can be partitioned into $k\le 3$ sets $P_1,\ldots,P_k$ such that for each $i\in\{1,\ldots,k\}$,
		\begin{compactenum}
			\item $F[P_i]$ is a path; and
			\item $|V(P_i)\cap L_j| \le 3$ for every integer $j\ge 0$.
		\end{compactenum}
		\item Let $N$ and $N'$ be the subgraphs of $G$ and $G'$ consisting only of those edges and vertices contained in the closure of $F$.
	\end{compactenum}
	Then $N$ has an $H$-partition $\PP=\{S_x: x\in V(H)\}$ such that:
	\begin{compactenum}[(i)]
		\item for each $i\in\{1,\ldots,k\}$, there exists $x_i\in V(H)$ such that $P_i=S_{x_i}$;
        \item For each $x\in V(H)$, $S_x=Y_x\cup X_x$ where $|X_x|\le 1$ and $|Y_x\cap L_j|\le 3$ for each $j\ge 0$;
        \item $H$ has a plane embedding in which $x_1,\ldots,x_k$ appear on the outer face;
		\item $\tw(H)\le 3$.
	\end{compactenum}
\end{lem}

@@

  The proof is by induction on the number of inner faces of $N'$.  A particular easy case occurs when $V(N')=V(F)$, for which we can simply take $\PP:=\{P_1,\ldots,P_k\}$ and verify that the preconditions of the lemma (the `setup') ensure that $\PP$ satisfies the requirements (i)--(iv):
  \begin{compactenum}[(i)]
    \item By definition, each of $P_1,\ldots,P_k$ is a part of $\mathcal{P}$.
    \item For each $x\in V(H)$, let $Y_x:=S_x$ and $X_x=\emptyset$.  Then 4(a) ensures that $|Y_x\cap L_j|\le 3$ for each integer $j\ge 0$.
    \item $H$ has $k\le 3$ vertices and any graph on $k\le 3$ vertices has a plane embedding with all its vertices on the outer face.
    \item The trivial tree decomposition of $H$ that places $x_1,\ldots,x_k$ in one bag has width $k-1\le 2\le 3$.
  \end{compactenum}
  From this point on, we assume that $N$ has at least one inner vertex.

  The $H$-partition $\mathcal{P}$ will include $P_1$, $P_2$ , $P_3$ and a (possibly empty) part $S$ that is defined by three vertices $v_1v_2v_3$ that belong to a common inner face $\tau$ of $N'$.  Depending on the value of $k$, there are two cases to consider:
  \begin{compactenum}
    \item If $k\in\{1,2\}$ then we let $\tau:=v_1,\ldots,v_p$ be an inner face of $N'$ that contains at least one edge of $F$ on its boundary.

@@

  We can now define the vertices in $S$.  For each $i\in\{1,2,3\}$, let $Q_i$ be the shortest path, in $T$, from $v_i$ to a vertex in $V(F)$.  Let $\overline{S}$ denote the subgraph of $N'$ consisting of vertices and edges of $Q_1$, $Q_2$, $Q_3$, and $\tau$.  Let $S:=V(\overline{S})\setminus V(F)$, let $Y:=(V(Q_1)\cup V(Q_2)\cup V(Q_3))\cap S$, and let $X:=S\setminus Y$.
  Observe that, for each $i\in\{1,2,3\}$ and each $j\ge 0$, $|V(Q_i)\cap L_j|\le 1$.  Therefore $|Y\cap L_j|\le 3$ for each integer $j\ge 0$. Furthermore, $|X|\le p-3\le 1$, since $v_1,v_2,v_3\in Y\cup V(N)$. Therefore, taking $Y:=S\cup\{v\}\setminus V(F)$ as a part in the partition $\PP$ will not violate condition~(ii).

@@

  Let $F_1,\ldots,F_m,\tau$ be the inner faces of $M$.   By our choice of $\tau$, the graph $M$ is $2$-connected.  Furthermore, $V(F_i)$ can be partitioned into at most three sets $P_{i,1}$, $P_{i,2}$, and $P_{i,3}$ where $P_{i,1}\subset Y$, $P_{i,2}\subseteq P_b$, and $P_{i,3}\subseteq P_c$ for some $b,c\in\{1,\ldots,k\}$.

  In the following, in order to avoid introducing even more notation or constantly adding the modifier ``when it exists'', we will assume that $V(F_i)\cap Y\neq\emptyset$ and that there are exactly two distinct $b,c\in\{1,\ldots,k\}$ such that $V(F_i)\cap P_b\neq\emptyset$ and $V(F_i)\cap P_c\neq\emptyset$.  When this assumption does not hold, exactly the same reasoning can be used, while omitting one or more of $P_{i,1}$, $P_{i,2}$ or $P_{i,3}$.

  Since $M$ is $2$-connected and $\overline{Y}$ is connected, $F_i[P_{i,j}]$ is a path, for each $j\in\{1,2,3\}$.  Let $N_i$ and $N_i'$ be the subgraphs of $N$ and $N'$ contained in the closure of $F_i$.  Every inner face of $N_i'$ is an inner face of $N'$ and $\tau$ is not an inner face of $N_i'$.  Therefore $N_i'$ has fewer inner faces than $N'$.  Therefore, we can apply induction using the cycle $F_i$ and the partition $P_{i,1},P_{i,2},P_{i,3}$ of $V(F_i)$.  The result of the induction is a $H_i$-partition $\mathcal{P}_i$ of the subgraph $N_i$ of $N$ contained in the closure of $F_i$.
  % The partition $\mathcal{P}_i$ contains each of $P_{i,1}$, $P_{i,2}$ and $P_{i,3}$.
  We can now define our desired partition
  \[
      \mathcal{P}:=\{Y, P_1,\ldots,P_k\} \cup \bigcup_{i=1}^m \{P\in\mathcal{P}_i: P\cap V(M)=\emptyset\} \enspace .
  \]
  It is straightforward to verify that $\mathcal{P}$ is indeed a partition of $V(N)$ and, by definition this is an $H$-partition of $N$ for the graph $H:=N/\mathcal{P}$.  All that remains is to verify that $\mathcal{P}$ satisfies conditions~(i)--(iv).
  \begin{compactenum}[(i)]
    \item By construction $\mathcal{P}$ contains parts $P_1,\ldots,P_k$.

    \item  By Assumption~4(b) none of the parts $P_1,P_2,P_3$ violate condition~(ii).  We have already argued that the part $S$ does not violate condition~(ii).  The inductive hypothesis ensures that each of the remaining parts satisfies condition~(ii).

    \item We only consider the (most difficult) case where $k=3$, the cases $\in\{1,2\}$ are handled easily. Let $x_0$ be the vertex of $H$ obtained from contracting $Y$ and, for each $j\in\{1,\ldots,3\}$, let $x_j$ denote the vertex of $H$ obtained from contracting $P_j$.  In this case the vertices $x_0,\ldots,x_3$ form a clique in $H$, which we embed with $x_1,\ldots,x_3$ on the outer face, as required.

    We claim that, for each distinct $a,b\in \{1,2,3\}$ there is at most one face $F_i$ such that $V(F_i)\cap S_{x_j}\neq\emptyset$ for each $j\in\{0,a,b\}$.  Suppose to the contrary that there are two such faces $F_{i_1}$ and $F_{i_2}$.  Create a graph $M^+$ from $M$ by placing a vertex $v_{i_b}$ inside $F_{i_b}$ and adjacent to each vertex in $V(F_{i_b})$ for each $b\in\{1,2\}$.  Since $M$ is planar and $F_{i_1}$ and $F_{i_2}$ are distinct faces of $M$, $M^+$ is planar.  However, contracting each of $S_{x_0},\ldots,S_{x_3}$ produces a graph isomorphic to the complete bipartite graph $K_{3,3}$, a contradiction.

    Therefore, by the inductive hypothesis, we can embed each such graph $H_i :=N_i/\mathcal{P}_i$ so that the three vertices on its outer face correspond to the appropriate inner face of the clique on $x_0,\ldots,x_3$.  Any remaining graphs $H_i$ have outer faces with at most two vertices on their outer faces corresponding to at most two vertices of $x_0,\ldots,x_3$.  These can easily be embedded, one a time, in some inner face that has the edge $x_ix_j$ on its boundary.

@@

    The tree $T_0$ has a node $y_0$ with $B_{y_0}:=\{x_0,\ldots,x_k\}$.
    % For each $i\in\{1,\ldots,m\}$, let $N_i$ denote the subgraph of $N$ contained in $F_i$.
    The inductive hypothesis ensures that $H_i:=N_i/\mathcal{P}_i$ has a tree decomposition $(B_z:z\in V(T_i))$ of width at most $3$.  Let $a_i,b_i,c_i\in V(H_i)$ be the vertices obtained by contracting $P_{i,1}$, $P_{i,2}$, and $P_{i,3}$, respectively. Since $a_i,b_i,c_i$ form a clique in $H_i$, some bag $B_{z_i}$ contains $a_i$, $b_i$, and $c_i$.

    Recall that $P_{i,1}:=V(N_i)\cap Y$, $P_{i,2}:=V(N_i)\cap P_b$ and $P_{i,3}:=V(N_i)\cap P_c$ for some $b,c\in\{1,\ldots,k\}$.  For each bag $B_z$, $z\in V(T_i)$, replace each occurrence of $a_i$ by $x_0$, replace each occurrence of $b_i$ by $x_b$ and replace each occurrence of $c_i$ by $x_c$.  These replacements do not increase the size of any bag and they result in a tree decomposition of the graph obtained from $H_i$ by renaming the vertices $a_i$, $b_i$, and $c_i$ with the names $x_0$, $x_a$, and $x_b$. Now add an edge that joins $y_0$ to $z_i$.

    Using the fact that each component of $G-V(M)$ is contained in a single face of $M$, it is now straightforward to verify that, by doing this for each $i\in\{1,\ldots,m\}$ we obtain a width-$3$ tree decomposition of $N$.  \qedhere
  \end{compactenum}

@@

	Let $G'$ be the plane graph obtained by removing each pair of crossing edges from $\overline{G}$ and let $T$ be a BFS tree of $G'$ rooted at $r$ and let $\langle L_0,L_1,\ldots,\rangle$ be the resulting BFS layering.  Now apply \cref{induction} with $G=\overline{G}$, $G'$, $F$, and $P_i=\{v_i\}$ for each $i\in\{1,2,3\}$.  This gives an $H$-partition $\mathcal{P}:=\{S_x\cup\{v_x\}:x\in V(H)\}$ of $\overline{G}-\{r\}\supseteq G$ in which $H$ is planar and has treewidth at most 3.

	Use the layering $\mathcal{L}=\langle L_0',L_1'\ldots\rangle$ where $L_i'=L_{2i}\cup L_{2i+1}$ for each integer $i\ge 0$. That this is a layering of $G$ follows from the fact that $\dist_{G'}(v,w)\le 2$ for each edge $vw\in E(G)$.  Since $|L_j\cap Y_x|\le 3$ for each integer $j\ge 0$, $|L_i'\cap Y_x|\le 6$ and therefore $|L_i'\cap S_x|\le 7$ for each integer $i\ge 0$ and each $x\in V(H)$. The result follows from \cref{PartitionProduct}.
\end{proof}
